Formalizing Commutative Algebra in Coq: \\ Nakayama's Lemma\thanks{Source code for this work is available on the following site: \url{https://github.com/ku-sldg/algebra}}

### 2022-11-05 nakayama/paper.tex
Edits

\vspace{-.3cm}

@@

Our goal is to newly formalize theoretical, rather than computational, commutative algebra in Coq. 
We formally prove \emph{Nakayama's lemma} \cite{nakayama-1951, azumaya}, an essential result in the field. 
In doing so, we formalize algebraic structures that are fundamental to higher-level algebra, such as 
\emph{local rings} and \emph{(finitely generated) modules over commutative rings}. 
Rather than build upon some of the basic objects from abstract algebra, such as groups and rings, that have been formalized in Coq, e.g., in the  
\emph{Mathematical Components Library} 
\cite{mathcomp}, we start from scratch. \todo{To do: Verify whether MathComp only formalized finite algebraic structures}
 The theory, including the formalization of all algebraic structures, is approximately  $100$ kB and \textcolor{red}{??}\todo{To do:  Add number} lines of code.

The notion of a module over a ring is an extension of the linear-algebraic notion of a vector space over a field, ubiquitous in mathematics and its applications. 
Less frequently referred to as the \emph{Krull-Azumaya theorem}\footnote{Hideyuki Matsumura explains in his text \emph{Commutative Algebra} \cite{matsumura}:  ``{This simple but important lemma is due to T.\ Nakayama, G.\ Azumaya
and W.\ Krull. Priority is obscure, and although it is usually called the Lemma of Nakayama, late Prof.\  Nakayama did not like the name.''}}\,\cite{nagata}, Nakayama's lemma 
describes one way that a finitely generated module over an arbitrary commutative ring acts like a vector space over a field.  
True to the convention that ``lemma'' often refers to a result serving as a stepping stone toward another goal, Nakayama's lemma is applied widely throughout the field, and the result is typically introduced in a first graduate course in commutative algebra \cite{atiyah-macdonald, matsumura, eisenbud}.

@@

When $R=F$ is a field, its unique maximal ideal is the zero ideal, and given any vector space $M=V$ over $F$, the only linear combination of vectors with coefficients in the zero ideal is the zero vector. Hence in this special case, the hypothesis that  $M = \mathfrak{m} M$ is equivalent to the conclusion that $M=0$. 
Hence Nakayama's lemma describes one way that finitely generated modules over commutative local rings are similar to vector spaces.

In general, the quotient $R/\mathfrak{m}$ of a local ring modulo its maximal ideal $\mathfrak{m}$ is a field, and the quotient of a module $M$ modulo the submodule $\mathfrak{m}M$ is an $R/\mathfrak{m}$-module, i.e., $M/\mathfrak{m}M$ is a vector space over $R/\mathfrak{m}$. Nakayama's lemma implies that if $M$ is finitely generated, then bases for $M/\mathfrak{m}M$ corresponds, via lifting, to minimal sets of generators of 
$M$.

@@

We start by describing the process of formalization of algebraic structures, and with that in hand, then move on to the formal proof of Nakayama's lemma.

@@

\begin{minted}{coq}
  Infix "==" := equiv (at level 60, no associativity).
  Class Semigroup := {
    semigroup_assoc:
      forall (a b c: Carrier),
        a <o> b <o> c == a <o> (b <o> c);
  }.
  Class Monoid := {
    monoid_semigroup :> Semigroup equiv op;
    monoid_ident_l:
      forall (a: Carrier), ident <o> a == a;
    monoid_ident_r:
      forall (a: Carrier), a <o> ident == a;
  }.
  Class Group := {
    group_monoid :> Monoid equiv op ident;
    group_inv_l:
      forall (a: Carrier), inv a <o> a == ident;
    group_inv_r:
      forall (a: Carrier), a <o> inv a == ident;
  }.
\end{minted}
The lines of the form, e.g.,  ``\verb|monoid_semigroup :> Semigroup equiv op;|'' simply coerce
the monoid typeclass into a semigroup.

%While we found later on that we did not need quotients, 
While in the end, our formal proof did not call upon quotients of algebraic structures, quotient rings and quotient modules are fundamental to commutative algebra, and one can use them to construct alternate proofs of Nakayama's lemma.
It is worth pointing out that we have formalized quotients of algebraic objects  in Coq using typeclasses, which appear to work rather nicely.

@@

\begin{minted}{coq}
  Definition left_congru (a b: Carrier) :=
    P (inv a <o> b).
  Definition right_congru (a b: Carrier) :=
    P (a <o> inv b).
\end{minted}
% \end{verbatim}
When these two relations coincide, then we can prove that this common equivalence
relation actually preserves the group structure. Subgroups which have this
property are called \emph{normal subgroups}.
% \begin{verbatim}
\begin{minted}{coq}
  Let normal_subgroup_congru_coincide :=
    forall (a b: Carrier),
      left_congru op inv P a b <->
      right_congru op inv P a b.

@@

Moving onward, we formalized the structure of a ring, which has two binary
operations:
addition, which must be commutative, and multiplication, which need not be commutative in general.
Next we defined the structure of a commutative ring, further requiring commutativity of multiplication, as well as a multiplicative identity. 
At this point,, we define {ideals} of commutative rings as normal subgroups of the ring under
addition that satisfy the absorption property under multiplication, i.e., \(r a\) is in
the ideal for every element \(a\) of the ideal, and every element \(r\) of the commutative ring.

Again, we defined quotient rings but found them unnecessary in the end. Note
that should an ideal \(I\) contain a unit, an element \(x\) with a
multiplicative inverse \(x^{-1}\), then \(I\) would be the entire ring by the
absorbing property: given any element \(a\) of the ring, then \(a x^{-1}\) is
also in the ring, and \(a x^{-1} x = a 1 = a\in I\). Maximal ideals were
defined next. These are ideals that are proper subsets of the ring while having
no larger ideal except for the ring itself. Below is the definition in Coq,
which uses \texttt{P} as the predicate for the ideal.
% \begin{verbatim}
\begin{minted}{coq}
  Definition maximal_ideal :=
    exists (r: Carrier), (not (P r) /\
      forall (Q: Carrier -> Prop)
          (Q_proper: Proper (equiv ==> iff) Q)
          (Q_ideal: Ideal add zero minus mul Q),
        (forall (r: Carrier), P r -> Q r) ->
        (forall (r: Carrier), Q r) \/
          (forall (r: Carrier), Q r -> P r)).
\end{minted}
% \end{verbatim}
Almost by definition, maximal ideals can contain no units, otherwise they would
fail to be a proper subset of the ring. We could then define a local ring,
which is a ring with a single maximal ideal.
% \begin{verbatim}
\begin{minted}{coq}
  Definition local_ring :=
    exists (P: Carrier -> Prop)
        (P_proper: Proper (equiv ==> iff) P)
        (P_ideal: Ideal add zero minus mul P),
      maximal_ideal P /\
      (forall (Q: Carrier -> Prop)
          (Q_proper: Proper (equiv ==> iff) Q)
          (Q_ideal: Ideal add zero minus mul Q),
        maximal_ideal Q -> forall (r: Carrier), P r <-> Q r).
\end{minted}
% \end{verbatim}

@@

The next structure defined were modules over rings which generalize vector
spaces over fields. We needed to capture the notion of linear combinations of
coefficients and module vectors. This was done by dependently typed vectors,
i.e. lists parameterized by their length. Because there is an overload of the
term ``vector'', we will use that term to refer to module vectors, and use the
term ``list'' to mean length parameterized lists. As we don't use the simpler
kind of lists, this avoids any name collisions. A finitely generated module is
like the vector space \(\mathbf{R}^{n}\) in that there are finitely many
vectors which can generate all other vectors, for \(\mathbf{R}^{n}\) one such
collection of generators are \(\mathbf{e}_{1} = \begin{pmatrix} 1 & 0 & 0 &
  \cdots & 0\end{pmatrix}^{T}\), \(\mathbf{e}_{2} = \begin{pmatrix} 0 & 1 & 0 &
  \cdots & 0\end{pmatrix}^{T}\), \textellipsis, \(\mathbf{e}_{n} =
  \begin{pmatrix} 0 & 0 & 0 & \cdots & 1\end{pmatrix}^{T}\). In our code,
\texttt{M} is the type of module elements, \texttt{R} is the type of ring
elements which act as coefficients, and \texttt{t A n} is a list whose elements
are of type \texttt{A} and whose length is \texttt{n}.
% \begin{verbatim}
\begin{minted}{coq}
  Definition finitely_generated {n: nat}(basis: t M n) :=
    forall (vector: M),
      exists (coeffs: t R n),
        vector =M= linear_combin coeffs basis.
\end{minted}
% \end{verbatim}

Next, we defined the product of an ideal \(I\) and a module \(M\), denoted as
\(I M\), which is defined to be all linear combinations of vectors from \(M\)
and coefficients from \(I\). This can be easily shown to be a submodule of
\(M\). We represented this in Coq as a predicate over \(M\).
% \begin{verbatim}
\begin{minted}{coq}
  Context (P: R -> Prop).
  Context {P_proper: Proper (Requiv ==> iff) P}.
  Context {P_ideal: Ideal Radd Rzero Rminus Rmul P}.

  Definition ideal_module (x: M): Prop :=
    exists (n: nat)(coeffs: t R n)(vectors: t M n),
      Forall P coeffs /\
      x =M= linear_combin Madd Mzero action coeffs vectors.
\end{minted}
% \end{verbatim}
The \verb|Forall P coeffs| ensures that every element of the coefficient list
\texttt{coeffs} satisfies the predicate \texttt{P}. From here, we can move to
stating Nakayama's lemma.

@@

\begin{theorem}[Nakayama's lemma]
  Let \(M\) be a finitely generated module over a local ring \(R\) with maximal
  ideal \(\mathfrak{m}\). If \(\mathfrak{m} M = M\), then \(M = 0\), which is
  to say that \(M\) is the zero module which consists of exactly one element,
  that being the zero vector.
\end{theorem}

@@

For the proof of Nakayama's lemma, it too follows the standard, informal
mathematical argument. Do induction on the number of vectors needed to generate
the module \(M\). The base case where \(M\) is generated by 0 vectors is by
definition true since an empty linear combination is conventionally taken to be
the zero vector. The inductive case where \(M\) is generated by the vectors
\(b_{1}, \dots, b_{m}, b_{m+1}\). By assumption \(M = \mathfrak{m} M\) with
\(b_{1}\in M\). Then \(b_{1}\in\mathfrak{m} M\) meaning that by our lemma,
there exists a linear combination for \(b_{1}\), say
\[b_{1} = u_{1} b_{1} + \cdots + u_{m+1} b_{m+1}\]
for some \(u_{1}, \dots, u_{m+1}\in\mathfrak{m}\). Collect the \(b_{1}\) terms
on the left-hand side of the equation to get that
\[(1 - u_{1}) b_{1} = u_{2} b_{2} + \cdots + u_{m+1} b_{m+1}\text{.}\]
As mentioned when defining maximal ideals, \(\mathfrak{m}\) can only contain
non-units. Namely, \(u_{1}\) must be a non-unit. Then we had already proven
that \(1 - u_{1}\) is a unit as we are in a local ring. Let \(v_{1}\) be the
multiplicative inverse of \(1 - u_{1}\), and multiply it on both sides of the
equation.
\begin{align*}
  v_{1} (1 - u_{1}) b_{1}
    & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
  1 b_{1} & = \\
  b_{1} & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
\end{align*}
We have thus found that one of the basis vectors is not needed to generate this
module. Using the induction hypothesis, we have that \(M = 0\). Showing that
the induction hypothesis holds in Coq takes more work than in an informal
proof, but this extra work is just a lot of bookkeeping.

>>>>

\vspace{-1cm}

@@

Our goal is to newly formalize theoretical, rather than computational, commutative algebra in Coq. 
We formally prove \emph{Nakayama's Lemma} \cite{nakayama-1951, azumaya}, an essential result in the field. 
In doing so, we formalize algebraic structures that are fundamental to higher-level algebra, such as 
\emph{local rings} and \emph{(finitely generated) modules over commutative rings}. 
Rather than build upon some of the basic objects from abstract algebra, such as groups and rings, that have been formalized in Coq, e.g., in the  
\emph{Mathematical Components Library} 
\cite{mathcomp}, we start from scratch. \todo{To do: Verify whether MathComp only formalized finite algebraic structures}
 The theory, including the formalization of all algebraic structures, is approximately  $100$ kB and \textcolor{red}{??}\todo{To do:  Add number} lines of code.

The notion of a module over a ring is an extension of the linear-algebraic notion of a vector space over a field, ubiquitous in mathematics and its applications. 
Less frequently referred to as the \emph{Krull-Azumaya theorem}\footnote{Hideyuki Matsumura explains in his text \emph{Commutative Algebra} \cite{matsumura}:  ``{This simple but important lemma is due to T.\ Nakayama, G.\ Azumaya
and W.\ Krull. Priority is obscure, and although it is usually called the Lemma of Nakayama, late Prof.\  Nakayama did not like the name.''}}\,\cite{nagata}, Nakayama's Lemma 
describes one way that a finitely generated module over an arbitrary commutative ring acts like a vector space over a field.  
True to the convention that ``lemma'' often refers to a result serving as a stepping stone toward another goal, Nakayama's Lemma is applied widely throughout the field, and the result is typically introduced in a first graduate course in commutative algebra \cite{atiyah-macdonald, matsumura, eisenbud}.

@@

When $R=F$ is a field, its unique maximal ideal is the zero ideal, and given any vector space $M=V$ over $F$, the only linear combination of vectors with coefficients in the zero ideal is the zero vector. Hence in this special case, the hypothesis that  $M = \mathfrak{m} M$ is equivalent to the conclusion that $M=0$. 
Hence Nakayama's Lemma describes one way that finitely generated modules over commutative local rings are similar to vector spaces.

In general, the quotient $R/\mathfrak{m}$ of a local ring modulo its maximal ideal $\mathfrak{m}$ is a field, and the quotient of a module $M$ modulo the submodule $\mathfrak{m}M$ is an $R/\mathfrak{m}$-module, i.e., $M/\mathfrak{m}M$ is a vector space over $R/\mathfrak{m}$. Nakayama's Lemma implies that if $M$ is finitely generated, then bases for $M/\mathfrak{m}M$ corresponds, via lifting, to minimal sets of generators of 
$M$.

@@

We start by describing the process of formalization of algebraic structures, and with that in hand, then move on to the formal proof of Nakayama's Lemma.

@@

\begin{minted}{coq}
  Infix "==" := equiv (at level 60, no associativity).
  Class Semigroup := {
    semigroup_assoc:
      forall (a b c: Carrier),
        a <o> b <o> c == a <o> (b <o> c);
  }.
  Class Monoid := {
    monoid_semigroup :> Semigroup equiv op;
    monoid_ident_l:
      forall (a: Carrier), ident <o> a == a;
    monoid_ident_r:
      forall (a: Carrier), a <o> ident == a;
  }.
  Class Group := {
    group_monoid :> Monoid equiv op ident;
    group_inv_l:
      forall (a: Carrier), inv a <o> a == ident;
    group_inv_r:
      forall (a: Carrier), a <o> inv a == ident;
  }.
\end{minted}
The lines such as  ``\verb|monoid_semigroup :> Semigroup equiv op;|'' simply coerce
the monoid typeclass into a semigroup.

%While we found later on that we did not need quotients, 
While in the end, our formal proof did not call upon quotients of algebraic structures, quotient rings and quotient modules are fundamental to commutative algebra, and one can use them to construct alternate proofs of Nakayama's Lemma.
It is worth pointing out that we have formalized quotients of algebraic objects  in Coq using typeclasses, which appear to work rather nicely.

@@

\begin{minted}{coq}
  Definition left_congru (a b: Carrier) :=
    P (inv a <o> b).
  Definition right_congru (a b: Carrier) :=
    P (a <o> inv b).
\end{minted}

When these two relations coincide, then we can prove that this common equivalence
relation actually preserves the group structure. Subgroups which have this
property are called \emph{normal subgroups}.

\begin{minted}{coq}
  Let normal_subgroup_congru_coincide :=
    forall (a b: Carrier),
      left_congru op inv P a b <->
      right_congru op inv P a b.

@@

Moving onward, we formalized the structure of a ring, which has two binary
operations:
addition, which must be commutative, and multiplication, which need not be commutative in general.
Next we defined the structure of a commutative ring, further requiring commutativity of multiplication, as well as a multiplicative identity. 
At this point, we formalized the notion of an ideal of a commutative rings as a normal subgroup  \emily{Emily:  Drew, this is what I read when I mentioned  ideals as normal subgroups.  We should check if this should be rewritten}
of the ring under
addition that satisfy the absorption property under multiplication, i.e., \(r a\) is in
the ideal for every element \(a\) of the ideal, and every element \(r\) of the commutative ring.
We also defined quotient rings.

Note
that should an ideal \(I\) contain a unit, an element \(x\) with a
multiplicative inverse \(x^{-1}\), then \(I\) would be the entire ring by the
absorbing property: given any element \(a\) of the ring, then \(a x^{-1}\) is
also in the ring, and \(a x^{-1} x = a 1 = a\in I\). 
\todo{To do: Check whether we need this paragraph}

Next, we formally defined maximal ideals.  Recall that a maximal ideal is a proper ideal that are maximal with respect to inclusion, i.e., if it is strictly contained in another ideal, this larger ideal must be the entire ring. 
Below is the definition in Coq,
which uses \texttt{P} as the predicate for the ideal.

\begin{minted}{coq}
  Definition maximal_ideal :=
    exists (r: Carrier), (not (P r) /\
      forall (Q: Carrier -> Prop)
          (Q_proper: Proper (equiv ==> iff) Q)
          (Q_ideal: Ideal add zero minus mul Q),
        (forall (r: Carrier), P r -> Q r) ->
        (forall (r: Carrier), Q r) \/
          (forall (r: Carrier), Q r -> P r)).
\end{minted}

Almost by definition, maximal ideals can contain no units, otherwise they would
fail to be a proper subset of the ring. \todo{To do:  Similar to above, do we need the first sentence here?}
 Using the definition of a maximal ideal, we could then define a local ring, i.e., a commutative ring with a single maximal ideal.

 \begin{minted}{coq}
  Definition local_ring :=
    exists (P: Carrier -> Prop)
        (P_proper: Proper (equiv ==> iff) P)
        (P_ideal: Ideal add zero minus mul P),
      maximal_ideal P /\
      (forall (Q: Carrier -> Prop)
          (Q_proper: Proper (equiv ==> iff) Q)
          (Q_ideal: Ideal add zero minus mul Q),
        maximal_ideal Q -> forall (r: Carrier), P r <-> Q r).
\end{minted}

@@

The next structure defined was a module over a commutative ring, the commutative-algebraic generalization of the notion of a vector
spaces over a fields,.  Here we needed to capture the notion of linear combinations of
coefficients and module vectors. This was done by dependently typed vectors,
i.e. lists parameterized by their length. Because there is an overload of the
term ``vector'', we will use that term to refer to module vectors, and use the
term ``list'' to mean length parameterized lists. As we don't use the simpler
kind of lists, this avoids any name collisions. A finitely generated module is
like the vector space \(\mathbf{R}^{n}\) in that there are finitely many
vectors which can generate all other vectors, for \(\mathbf{R}^{n}\) one such
collection of generators are \(\mathbf{e}_{1} = \begin{pmatrix} 1 & 0 & 0 &
  \cdots & 0\end{pmatrix}^{T}\), \(\mathbf{e}_{2} = \begin{pmatrix} 0 & 1 & 0 &
  \cdots & 0\end{pmatrix}^{T}\), \textellipsis, \(\mathbf{e}_{n} =
  \begin{pmatrix} 0 & 0 & 0 & \cdots & 1\end{pmatrix}^{T}\). In our code,
\texttt{M} is the type of module elements, \texttt{R} is the type of ring
elements which act as coefficients, and \texttt{t A n} is a list whose elements
are of type \texttt{A} and whose length is \texttt{n}.
% \begin{verbatim}
\begin{minted}{coq}
  Definition finitely_generated {n: nat}(basis: t M n) :=
    forall (vector: M),
      exists (coeffs: t R n),
        vector =M= linear_combin coeffs basis.
\end{minted}
% \end{verbatim}

Next, given an ideal \(I\) of a commutative ring $R$ and and an $R$-module \(M\), 
we defined the submodule $IM$, the set consisting of all scalar combinations of elements of \(M\) whose coefficients are in  \(I\). This can be easily shown to be a submodule of
\(M\). We represented this in Coq as a predicate over \(M\).
% \begin{verbatim}
\begin{minted}{coq}
  Context (P: R -> Prop).
  Context {P_proper: Proper (Requiv ==> iff) P}.
  Context {P_ideal: Ideal Radd Rzero Rminus Rmul P}.

  Definition ideal_module (x: M): Prop :=
    exists (n: nat)(coeffs: t R n)(vectors: t M n),
      Forall P coeffs /\
      x =M= linear_combin Madd Mzero action coeffs vectors.
\end{minted}
% \end{verbatim}
The use of  ``\verb|Forall P coeffs|'' ensures that every element of the coefficient list
\texttt{coeffs} satisfies the predicate \texttt{P}.

@@

Finally, we turn to our formal proof of Nakayama's Lemma, our ultimate goal.

\begin{nak*}
Let $R$ be a commutative local ring, and let $\mathfrak{m}$ denote its maximal ideal. 
Suppose that $M$ is a finitely generated $R$-module. 
If $M = \mathfrak{m} M$, then $M = 0$, i.e., $M$ must be the $R$-module containing only one element, its identity as an additive abelian group. 
\end{nak*}

@@

For the proof of Nakayama's Lemma, it too follows the standard, informal
mathematical argument. Do induction on the number of vectors needed to generate
the module \(M\). The base case where \(M\) is generated by 0 vectors is by
definition true since an empty linear combination is conventionally taken to be
the zero vector. The inductive case where \(M\) is generated by the vectors
\(b_{1}, \dots, b_{m}, b_{m+1}\). By assumption \(M = \mathfrak{m} M\) with
\(b_{1}\in M\). Then \(b_{1}\in\mathfrak{m} M\) meaning that by our lemma,
there exists a linear combination for \(b_{1}\), say
\[b_{1} = u_{1} b_{1} + \cdots + u_{m+1} b_{m+1}\]
for some \(u_{1}, \dots, u_{m+1}\in\mathfrak{m}\). Collect the \(b_{1}\) terms
on the left-hand side of the equation to get that
\[(1 - u_{1}) b_{1} = u_{2} b_{2} + \cdots + u_{m+1} b_{m+1}\text{.}\]
As mentioned when defining maximal ideals, \(\mathfrak{m}\) can only contain
non-units. Namely, \(u_{1}\) must be a non-unit. Then we had already proven
that \(1 - u_{1}\) is a unit as we are in a local ring. Let \(v_{1}\) be the
multiplicative inverse of \(1 - u_{1}\), and multiply it on both sides of the
equation.
\begin{align*}
  v_{1} (1 - u_{1}) b_{1}
    & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
  1 b_{1} & = \\
  b_{1} & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
\end{align*}
We have thus found that one of the basis vectors is not needed to generate this
module. Using the induction hypothesis, we have that \(M = 0\). Showing that
the induction hypothesis holds in Coq takes more work than in an informal
proof, but this extra work is just a lot of bookkeeping.
